```{r setup, results="hide", echo=FALSE}
require("dplyr", warn.conflicts=FALSE, quietly = TRUE)
#options(knitr.package.echo=FALSE)
#options(knitr.package.dev=c("svg"))
```
# Exploring No Show data

This is an exploratory data analysis (EDA), of data descriving medical
appointments.

The objectives are: first, to visually present the structure of the
sample data; and second, to show how the appointments' features relate
to whether the patients will show up nor not.

```{r helper_functions}
const <- function (v = stop("v not supplied")) {
    tmp <- substitute(function() {
        return(v)
    })
    eval(tmp, parent.frame())
}
`%$%` <- function (f = stop("f not supplied"), v = stop("v not supplied")) {
    tmp <- substitute((f(v)))
    eval(tmp, parent.frame())
}
```

```{r load_data, cache=TRUE}
#Loading the data
appData <-
	read.csv(
		 unz("data/noshowappointments.zip", "KaggleV2-May-2016.csv"),
		 header=TRUE)
```

## Medical Appointment Data.

The appointments' characteristics present in the data are:

```{r structure}
str(appData)
```

Some of them might require a short explanation:

* **ScheduledDay**: The day the appointment was scheduled.
* **AppointmentDay**: The day of the appointment.
* **Neighbourhood**: The Neighbourhood where the clinic is located.
* **Scholarship**: 1 if the person receives
  [aid from the government](https://en.wikipedia.org/wiki/Bolsa_Fam%C3%ADlia).
* **Handcap**: According to the source, A number that represents the person's
  disabilities.
* **SMS_received**: Whether the people received an SMS reminding them of the
  appointment. 
* **No.show**: The value "Yes", means that the person missed the appointment.
  The value "No", means that the person didn't miss the appointment.

### Computing Derived Features.

One of the factors that might be relevant to the show up ratio is the
length of time from the day the appointment was scheduled to the appointment
itself: 

**Wait** the number of days from the scheduling date to the appointment date.

```{r compute_wait, cache=TRUE}
schedDay <- const (
	sapply(X = (appData$ScheduledDay %>% as.character %>% strsplit("T")),
	       FUN=`[[`, 1) %>%
	strptime("%Y-%m-%d") %>% as.Date)


appDay <- const (
	sapply(X = (appData$AppointmentDay %>% as.character %>% strsplit("T")),
	       FUN=`[[`, 1) %>%
	strptime("%Y-%m-%d") %>% as.Date)

appData$Wait= (appDay() - schedDay()) %>% as.numeric #in days
appData$ScheduledDay = schedDay()
appData$AppointmentDay = appDay()
```

### Removing malformed records.

Some rows with negative `Wait` or `Age` where ignored when performing the
analysis.

```{r remove_negative_wait_and_age}
appData <- appData[appData$Wait>=0,]
appData <- appData[appData$Age>=0,]
```

## Exploration

This section contains the actual exploration of the data after having
established what features are present and removed the malformed records.

### Some numbers

The first set of questions are about quantities. 

```{r eda}
eda <- list()
eda$n_samples <- nrow(appData) #110,521
eda$n_patients <- length(unique(appData$PatientId)) #62,298
eda$n_appointments <- length(unique(appData$AppointmentID)) #one row per appointment
eda$n_neighbourhoods <- nlevels(appData$Neighbourhood) #81
eda$range_wait <- range( appData$Wait ) # Waits of up to 179 days
eda$range_scheduledDay <- range( appData$ScheduledDay ) #"2015-11-10" "2016-06-08"
eda$range_appointmentDay <- range( appData$AppointmentDay ) #"2016-04-29" "2016-06-08"
```

* How many appointments? `r format(eda$n_samples, big.mark=" ", scientific=FALSE)` appointments.
* How many patients? `r format(eda$n_patients , big.mark=" ", scientific=FALSE)` patients.
* Where do the appointments take place? In `r eda$n_neighbourhoods` different neighbourhoods.
* When were the appointments scheduled? From `r paste(as.character(eda$range_scheduledDay), collapse=" to ")`.
* When was the appointment? From `r paste(as.character(eda$range_appointmentDay), collapse=" to ")`.
* How many days did the patients had to wait? From `r paste(as.character(eda$range_wait), collapse=" to ")` days.

### The target composition.

Now a visualization of the sample's composition. Starting with the proportion
of missed appointments, which is the target variable of the study. The data
shows that most patients don't miss their appointments.

```{r no_show_distribution}
middle_points <- table(appData$No.show) %>% barplot(col=c(2,4))
proportions <- table(appData$No.show) %>% prop.table %>% c 
text(middle_points, y=10000, labels = Vectorize(function(v) { sprintf("%2.0f %%",v)
						})(proportions*100))
legend("topright", title="No.show", fill=c(2,4),
       legend=levels(appData$No.show))
title(main="Most patients don't miss their appointments")
```

### Features composition.

It is also interesting to visualize the relation between the features and the
target.

Two questions about the features are explored: what is the feature's
distribution in the data?; and, what is it relation with the target?

**Gender**

There are almost twice as many women than men.
Both genders are equally likely to miss their appointments.

```{r gender_distribution, fig.height=7}
par(mfrow=c(2,1))
tbl_gender <- table(appData$No.show, appData$Gender) 
prop_gender <- table(appData$Gender) %>% prop.table %>% c 
middle_points <- barplot(tbl_gender, xlab="Gender", ylab="count", col=c(2,4))
text(middle_points, y=10000, labels = Vectorize(function(v) { sprintf("%2.0f %%",v)
						})(prop_gender*100))
title(main="Most patients are female.")
legend("topright", title="No.show", fill=c(2,4),
       legend=levels(appData$No.show), cex=0.7)
tbl_gender %>% prop.table(2) %>% 
	barplot(xlab="Gender", ylab="Proportion", col=c(2,4))
title("Both genders are equally likely to miss the appointment")
```

**Patient's Age**

The patients' age distribution. The relation with whether they show up or don't
indicates that patients around the 15-20 age range are slightly more likely to
miss the appointment.

As expected, there are few samples for very old ages.

```{r hist_ages_1, fig.show="hide"}
hist_age<-hist(appData$Age)
```
```{r hist_ages_2, fig.height=7}
par(mfrow=c(2,1))
barplot(table(appData$No.show, cut(appData$Age, hist_age$breaks,
	include.lowest=TRUE)), xlab="Age Bracket", ylab="count", col=c(2,4))
title("Histogram of Age")
legend("topright", title="No.show", fill=c(2,4), legend=levels(appData$No.show))
barplot(table(appData$No.show, cut(appData$Age, hist_age$breaks)) %>% prop.table(2),
	xlab="Age Bracket", ylab="Proportion", col=c(2,4))
title("Proportion of No.show by age")
```

**Waiting Time**

Most appointments are same-day or for the next day. Those
appointments also have better turn up. Lower waiting times appear to improve the
show up ratio.  

*Warning*, the histograms below have non-uniform bin sizes.

```{r hist_wait, fig.show="hide"}
hist_wait <- hist(appData$Wait, breaks=Vectorize(function(e) {2**e})(seq(0,8))-1)
```

```{r hist_wait_2, fig.height=7}
par(mfrow=c(2,1))
tbl_wait <- table(appData$No.show, cut(appData$Wait, hist_wait$breaks,
				       include.lowest=TRUE))
barplot(tbl_wait, xlab="Wait Bracket", ylab="count", col=c(2,4))
legend("topright", title="No.show", fill=c(2,4),
       legend=levels(appData$No.show))
title("Histogram of Wait")
barplot(tbl_wait %>% prop.table(2), xlab="Wait",
			 ylab="Proportion of No.show", col=c(2,4))
title(main="Lower wait times have better turn up.")
```

**Distribution of Neighbourhoods**

There is considerable variance in the quantities of patients by neighbourhood,
but not so in the show up ratio.

```{r hist_neighbourhood, fig.height=7, cache=TRUE}
tbl_neigh <- table(appData$No.show, appData$Neighbourhood)
par(mfrow=c(2,1))
barplot(tbl_neigh, col=c(2,4),
	ylab="Count", xlab="Neighbourhood")
legend("topleft", title="No.show", fill=c(2,4),
       legend=levels(appData$No.show), cex=0.7, xpd=TRUE)
barplot(tbl_neigh %>% prop.table(2), col=c(2,4),
	ylab="Proportion", xlab="Neighbourhood")
title(main="Most locations have the same show up ratio")
```

**Distribution of other categorical variables**

Most notably, it appears that patients that received an SMS (`SMS_received == 1`)
where also less likely to show up.

This is due the fact that patients with same-day appointments don't receive a
SMS and are, simultaneously the more likely to show up (maybe they already at
the hospital).

```{r hists_categorical_variables, results="hide", fig.height=20}
par(oma = c(0,0,2,0))
hs <- c(0.75,rep(1,6))
hs <- hs / sum(hs)
layout(matrix(c(1,1,seq(2,13)),nrow = 7,ncol = 2,byrow = TRUE)
       ,heights = hs)
plot(1, type = "n", axes=FALSE, xlab="", ylab="")
legend("bottom", inset=0, title="No.show",
       pch=c(15,15,21), col=c(2,4),
       legend=c("No", "Yes"),
       , horiz=TRUE)
title(main="Histogram/Composition", sub="Maybe SMS = 1 means no SMS was sent.")

mapply(FUN=function(column,varname) {
	tbl_var <- table(appData$No.show, column)
	middle_points <- barplot(tbl_var, xlab=varname, ylab="Count", col=c(2,4))
	text(middle_points, y=40000, labels = Vectorize(function(v) { sprintf("%2.0f %%",v)
						})(100*(table(column) %>%
							prop.table %>% c)))
	barplot(prop.table(tbl_var,2), xlab=varname,
				 ylab="Proportion", col=c(2,4))
},appData[,8:13], names(appData)[8:13])
```

```{r same_day_sms}
with(appData %>% filter(Wait == 0), { table(SMS_received) })
```

Actually, if we only consider appointments that don't occur the same day,
patients who received a SMS a bit more likely to show up.

```{r not_same_day_sms}
with(appData %>% filter(Wait != 0), {
     par(mfrow=c(1,2), oma = c(0, 0, 2, 0))
     tbl_var <- table(No.show, SMS_received)
     middle_points <- barplot(tbl_var, xlab="SMS_received", ylab="Count", col=c(2,4))
     text(middle_points, y=40000, labels = Vectorize(function(v) { sprintf("%2.0f %%",v)
						})(100*(table(SMS_received) %>%
							prop.table %>% c)))
     barplot(prop.table(tbl_var,2), xlab="SMS_received",
	     ylab="Proportion", col=c(2,4))
     mtext("Sending SMS is useful when Wait > 0.", outer = TRUE, cex = 1.5)
})
```

## Some Characteristics

**Are there patients with more than one appointment on the same day?**

Yes, but most have only one.

```{r multiple_appointments}
ordAppData <- appData[order(appData$AppointmentDay),]
ordAppData %>% group_by(PatientId, AppointmentDay) %>%
	summarize(C=n()) %>% ungroup() %>% select(C) %>% `[[`(1) %>%
	table(dnn="Appointments per day")
```

**How does showing up for the prevoius appointment predicts showing up for the
next?**

Patients that showed up for their previous appointment are more
likely to show up again.

```{r previous_appointment, cache=TRUE}
last_outcome <- ordAppData %>% group_by(PatientId, AppointmentDay) %>%
	summarize(nYes=sum(No.show == "Yes"), nNo=sum(No.show == "No")) %>% ungroup %>%
	group_by(PatientId) %>%
	mutate(pnYes=lag(nYes), pnNo=lag(nNo)) %>% ungroup

ordAppDataWithLastOutcome <- inner_join(ordAppData, last_outcome, by= c("PatientId","AppointmentDay"))

nextVsPrev <- const %$% table(ordAppDataWithLastOutcome$No.show,
			      ordAppDataWithLastOutcome$pnYes > 0, useNA="ifany")

par(mfrow=c(1,2), oma = c(0, 0, 2, 0))
middle_points <- nextVsPrev() %>%
 	barplot(xlab="Previos No.show", ylab="count",
		 names.arg=c("No", "Yes", "NA"), col=c("red","blue"))
text(middle_points, y=20000, labels =
     Vectorize(function(v) { sprintf("%2.0f %%",v) })(
 	100*(nextVsPrev() %>% margin.table(2) %>% prop.table)))
legend(x=0.5,y=60000, title="No.show", fill=c("red","blue"),
       legend=levels(appData$No.show))
nextVsPrev() %>% prop.table(2) %>%
	barplot(xlab="Previos No.show", ylab="Proportion",
		names.arg=c("No", "Yes", "NA"), col=c("red","blue"))
mtext("Next No.show vs previous No.show.", outer = TRUE, cex = 1.5)
```

## [EXTRA]Gender Composition By Age

**Why is the proportion of women so large?**

```{r age_by_gender, results="hide", fig.height=7 }
par(mfrow=c(2,1))
table(appData$Gender, cut(appData$Age, hist_age$breaks , include.lowest=TRUE)) %>%
	barplot(xlab="Age Bracket", ylab="Count", col=c(2,4))
legend("topright", title="Legend",
       pch=c(15,15), col=c(2,4),
       legend=c("Female", "Male")) 
table(appData$Gender, cut(appData$Age, hist_age$breaks,
			  include.lowest=TRUE)) %>%
	prop.table(2) %>% barplot(xlab="Age Bracket", ylab="Proportion", col=c(2,4))
title(main="Women are the majority after the age of 16.")
```

There is a sharp drop in the proportion of men after the age of 16.

**License**

Released under [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/).
Attributed to Kaggle's user [JoniHoppen](https://www.kaggle.com/joniarroba)
([Twitter](https://twitter.com/jonihoppen),[LinkedIn](https://www.linkedin.com/in/joniarroba/)).
Changes where made to the data as shown in the code.
